---
title: "The Relations between Gas Mileage and Transmissions in Automobile"
author: "Taedong Yun"
date: "Saturday, April 25, 2015"
output: html_document
---

## Introduction and Explorative Data Analysis

Let us first explore the mtcars dataset.

```{r}
data(mtcars)
str(mtcars)
sum(is.na(mtcars)) # number of null values
nrow(mtcars) # total number of observations
mean(mtcars$mpg) # mean mpg
sum(mtcars$am) # number of cars with manual transmission
```

possible confounders: weight, horsepower or displacement or number of cylinders

nested likelihood ratio tests:
```{r}
mtcars$am <- factor(mtcars$am)
fit1 <- lm(mpg ~ am, data = mtcars)
fit2 <- update(fit1, mpg ~ am + hp)
fit3 <- update(fit1, mpg ~ am + hp + wt)
anova(fit1, fit2, fit3)
```

choose the third model

```{r}
fit <- lm(mpg ~ am + hp + wt, data = mtcars)
summary(fit)
```


* interpret coefficients
* exploratory data analysis
* fit multiple models and detail their strategy for model selection
* answer the questions of interest (or detail why the question is not answerable)
* residual plot and some diagnostics
* quantify the uncertainty in the conclusions and/or perform an inference
* report < 2 page, including appendix of figures < 5 page
* executive summary